Biodiversity improves tropical forest resilience

### 2019-01-11 Main Document.Rmd
# Draft 0.1.9014 * Submission

---
title: Biodiversity improves tropical forest resilience
author: Sylvain Schmitt$^{*,1,2}$, Isabelle Maréchaux$^3$, Jerome Chave $^4$, Fabian Fischer$^4$, Camille  Piponiot$^5$, Stéphane Traissac$^6$, Bruno Hérault$^{7,8}$
output:
  bookdown::pdf_document2:
    includes:
      before_body: ./template/doc_prefix.tex
      in_header: ./template/preamble.tex
    number_sections: no
    toc: no
    keep_tex: yes
linestretch: 1.5
csl: /home/sylvain/Documents/Bibliography/csl/joe.csl
bibliography: /home/sylvain/Documents/Bibliography/library.bib
link-citations: yes
---

@@

$^3$ AMAP, INRA, University of Montpellier, IRD, CIRAD, CNRS, , Montpellier, France

@@

$^5$ Université de Guyane, UMR EcoFoG (Agroparistech, CNRS, Inra, Université des
Antilles, Cirad), Kourou, French Guiana

@@

$^7$ Cirad, UMR EcoFoG (AgroParistech, CNRS, Inra, Université des Antilles, Université de la Guyane), Campus Agronomique, 97310 Kourou, French Guiana

$^8$ INPHB, Institut National Polytechnique Félix Houphouët-Boigny, BP 1093, Yamoussoukro, Ivory Coast

__$^*$ Corresponding author:__ sylvain.schmitt@agroparistech.fr

@@

1. Climate change and direct human activities are currently deeply modifying forest disturbance regimes, highlighting the critical need to better understand the drivers of forest resilience. In tropical forests, the role of biodiversity in ecosystem recovery and the underlying mechanisms remain under-documented. Studies relying on forest stand inventories often lack repeatability, rarely include disturbed forests, and temporal coverage remains too low to assess long-term effects. Forest simulators have thus proven useful to investigate the effect of biodiversity on forest functioning through virtual biodiversity experiments. However, the problem of biodiversity representation beyond a few plant functional types has long been an impediment in applying this approach to species-rich tropical forests, until now.
2. Using an individual-based and spatially-explicit model of tropical forest simulating high-diversity levels, we assessed *in silico* the effect of species and functional diversity on long-term forest recovery from disturbances of different levels. We used this approach to manipulate species number and identity across a large number of simulations, and measured the net effect of biodiversity on simulated forest recovery after disturbance. In order to further investigate the mechanisms underlying the effect of biodiversity, we partitioned ecosystem properties into complementarity and selection effects.
3.  We found that functional diversity improved tropical forest resilience from disturbance with underlying ecological mechanisms changing in dominance with time. Complementarity effect was dominant soon after disturbance and was progressively overtaken by a selection effect as more competitive species dominated the forest community, a pattern that increased with disturbance intensity.
4. *Synthesis*. We found that the mechanisms through which biodiversity influence forest functioning strongly depend on the current ecosystem state and its previous history of disturbances, shifting from dominant complementarity effect in recently disturbed ecosystems to selection effect in anciently disturbed systems. Our results suggest that pre-disturbance history is a key to our understanding of the ecological mechanisms that underlie the resilience of tropical forests and can help reconciling previous contrasted results obtained with limited snapshots of ecosystem state.

<!-- *The Abstract should be written to be understood in isolation from the rest of the paper and must not exceed 350 words. Using typically four simple, factual, numbered statements, it should: (i) give the conceptual context, (ii) state the methodological approach, (iii) report the main results and conclusions, and (iv) include a final point headed ‘Synthesis’, which sums up the paper’s key message in generic terms that can be understood by non-specialists, indicating clearly how this study has advanced ecological understanding.* -->

@@

Resilience; Recovery; Functional diversity; Tropical forest; Complementarity effect; Selection effect; Biodiversity-ecosystem functioning;

@@

Climate change and direct human activities are currently deeply modifying forest disturbance regimes [@Davidson2012; @Lewis2015], and while deforestation have been rightly the focus of worldwide attention, the consequences of human-induced disturbance on tropical forests has been less studied [@Putz2010; @Simula2009]. Degraded forests currently cover areas over ten times larger than deforested ones [@Herold2011], a majority of which is being unsustainably logged for timber production [@Blaser2011]. Forest degradation can trigger severe biodiversity loss [@Burivalova2014] and is estimated to account for 2.1 billion of tons of carbon dioxide emissions per year representing 25% of greenhouse gas emissions related to deforestation and forest degradation globally [@Pearson2017].  In addition to direct human impact on forests, climate change modifies regional climates with, for instance, projected increases in the frequency and intensity of convective storms in the Amazon having the potential to increase wind-related tree mortality [@Negron-Juarez2018]. Understanding the drivers of tropical forest resilience to disturbance is thus critical. Here we explored the role of biodiversity in tropical forest recovery after disturbance through a simulation approach.

Ecological resilience refers to the ecosystem ability to resist and to recover from disturbance [@Gunderson2010]. Following a disturbance, ecosystem recovery is determined by both exogenous (*e.g.* climate, disturbance features) and endogenous (*e.g.* forest structure and composition) factors. Logging intensity (the quantity of extracted timber and subsequent biomass loss) determines to a large extent the time-to-recovery of post-logging tree carbon, subsequent animal biodiversity loss, and shifts in tree species composition [@Burivalova2014; @DeAvila2015; @Rutishauser2015]. Among endogenous drivers, baseline floristic composition appears to be a key driver of forest recovery [@Herault2018], as species respond differently to disturbances depending on their ecological strategies and functional properties [@herault_functional_2011]. Beyond species identity, tree community diversity may also foster forest resilience to disturbances [@Cadotte2011; @Ives2007; @Loreau2013].

Two types of mechanisms can underlie the effect of biodiversity on ecosystem properties: complementarity and selection effects [@Loreau2001]. The complementarity effect results from resource partitioning and niche differentiation, and to facilitation, leading to a more efficient use of available resources at the community level. The selection effect describes how more diverse communities have a higher probability of including functionally efficient species. Moreover the sampling effect implies redundancy and a lower risk to lose important functional traits, further improving resilience of species rich-forest [@Hooper2005a; @Tilman2001]. Previous theoretical work [@Loreau2010], experiments [@Cardinale2007; @Cardinale2009; @Niklaus2017a; @Huang2018], and direct observations along natural gradients [@Chisholm2013; @Liang2016; @Paquette2011; @Poorter2017] suggest that both complementarity and selection mechanisms play a crucial role in ecosystem biomass stock and dynamics [@Cardinale2007; @Poorter2017; @VanderSande2017a]. The relative importance of each mechanism varies in time: for instance @Fargione2007 observed a shift from selection to complementarity effect over time in experimental grasslands, and @Huang2018 observed an increased complementarity over time with no selection effect in forest establishments experiments.

@@

In the present study, we take advantage of recent developments made in the forest simulator TROLL [@Chave1999], an individual-based and spatially-explicit model of tropical forest, that simulates high-diversity levels through a species-specific parameterization using plant functional traits [@Marechaux2017]. Consequently, TROLL offers the opportunity to study biodiversity-ecosystem functioning  relationships *in silico* in tropical forest communities. More specifically, we used a simulation approach to assess the effect of species and functional diversity on long-term forest recovery from disturbances of different levels. We used this approach to manipulate species number and identity across a large number of simulations, and measured the net effect of biodiversity on simulated forest recovery after disturbance. In order to further investigate the mechanisms underlying the effect of biodiversity, we partitioned ecosystem properties into complementarity and selection effects. Based on the general hypothesis of a positive relationships between biodiversity and productivity [@Liang2016], we hypothesized that more diverse forest would recover quicker to a disturbance event due to an increased productivity [but see @VanderSande2018 for no effect of species richness on productivity].

@@

TROLL is an individual-based and spatially explicit forest dynamic model. The model is described in detail in @Marechaux2017, and only summarized here. In TROLL the life cycle (recruitment, growth, seed production and death) is simulated for individual trees $>1\textrm{cm}$ diameter at breast height (dbh). Trees grow in a light environment computed within a $1\textrm{m}^3$ voxel grid. Seedlings below the $1\textrm{cm}$ size class are not modeled on an individual basis, but are part of a seed/seedlings pool. A maximum of one tree can establish per squaremeter. Once established, the tree is defined by age, dbh, height (h), crown radius (CR), crown depth (CD) and leaf area (LA). Tree geometry is calculated with allometric relationships. Additionally, each tree is flagged with a species label inherited from the parent tree through seedling recruitment. To each species label is associated a number of species-specific traits (Table \@ref(tab:traits)) from which ecophysiological processes are simulated using relationships drawn from the literature. The model thus represents tree biodiversity in a realistic way, where data requirement and model outputs parallel current trait collection efforts and field inventories.

Carbon assimilation is computed over half-hourly period of a representative day per month using the Farquhar, von Caemmerer, and Berry model of photosynthesis [@Farquhar1980].  Photosynthetic capacities, as well as leaf respiration rate, are computed from species-specific leaf concentration of nitrogen (N) and phosphorous (P), and leaf mass per area [LMA; @Atkin2015; @Domingues2010]. The net amount of assimilated carbon allocated to growth is then partitioned among the different plant organs using empirical factors, and the fraction allocated to woody growth is converted into increments of stem volume using species-specific wood density; while the fraction allocated to leaves is converted into leaf area increment using LMA. Leaf demography is simulated using species-specific leaf lifespan derived from LMA. Here we did not use the same relationship between LMA and leaf lifespan as in @Marechaux2017, as we found that it generated an underestimation of leaf lifespan for low LMA species. We thus designed a new leaf lifespan allometry, specific to tropical tree species, used throughout this study (see supplementary materials S1). Natural tree deaths are stochastic events and their rate is negatively correlated with wood density within TROLL, as observed pan-tropically [@kraft_functional_2010; @poorter_are_2008; @Wright2010] an also locally [@Aubry-Kientz2013]. Additionally, trees can die due to carbon starvation or treefalls. Below-ground processes, herbaceous plants, epiphytes and lianas are not simulated in this version of TROLL. The source code is written in C++ and version 2.3.1 used in this study are available at https://github.com/TROLL-code/TROLL. A further detailed description of the model is available in @Marechaux2017.

In order to have a general overview of disturbance effect, a generic disturbance module was developed for this study. At a given iteration $disturb_{iter}$, individual trees are randomly picked following a uniform law, targeting neither individuals nor species. This ensures that there are no confounding factors due to disturbance type specificity. Selected individuals are then removed without triggering a treefall to control disturbed biomass volume. The operation is repeated until the cumulative basal area of removed trees reaches targeted threshold ($disturb_{intensity}$).

@@

In order to assess the role of biodiversity on ecosystem recovery from disturbance, we  simulated forest trajectories for different levels of disturbance intensity ($disturb_{intensity}$) and tree community compositions. For each simulation, we first conducted a spin-up run, where we simulated forest regeneration from bare soil for 600 years (pre-disturbance step). We then simulated the disturbance described as above ($disturb_{iter}$), and finally let the simulated forest community freely recover from the disturbance during 600 years (post-disturbance step). Disturbance intensity ($disturb_{intensity}$) was fixed at 0% (control), 25%, 50% or 75% of basal area loss. For each level of disturbance, we performed 20 simulations differing in the original number of species (species richness $SR\{15,25,125\}$) and in floristic composition. For each level of species richness, we first created 1000 virtual communities by randomly picking the desired number of species among the regional species pool. Then of those 1000 communities, we selected 20 communities with a large range of convex hull volume $CVH$ [@Cornwell2006a] but with a community mean ($CM$) close to the one of the regional species pool ($CM_{LMA} = 115 \pm 19 ~\textrm{g.m}^{-2}$, $CM_{wsg} = 0.73 \pm 0.07~\textrm{g.cm}^{-3}$, $CM_{dmax} = 0.37 \pm 0.1~\textrm{m}$, $CM_{hmax} = 40 \pm 11~\textrm{m}$) in order to control for average functional composition. This led to 240 simulations varying in community composition and disturbance intensity ($3~SR*20~CHV*4~ levels~of~disturbance$). We focused on functional richness with convex hull volume $CVH$ and functional mean with community weighted mean $CWM$ to define simulation communities.

@@

where $X_D(t)$ and $X_C(t)$ are the values for the ecosystem property $X$ at time $t$ in the disturbed simulation ($disturb_{intensity} = 25\%, 50\%, 75\%$) and in the control simulation ($disturb_{intensity} = 0\%$), respectively. Note that variability among replicated simulations with the same set-up, due to demographic stochasticity in the model, is low enough to assume a given simulation is representative of the community properties and dynamics given an initial species pool and disturbance intensity [@Marechaux2017].

When $R(t)$ approaches $R_{eq}=1$, the disturbed and control systems are considered equivalent, indicating a perfect recovery. We then calculated the Euclidean distance to this state of perfect recovery as $d_{eq}(t) = \sqrt{(R_{eq}-R(t))^2}$ (Fig. \@ref(fig:IrGraphFinal)). Because we set up the intensity of disturbance, our analysis did not include any ecosystem resistance, i.e. the ecosystem capacity to absorb a disturbance event. Consequently, in this particular case, we used the inverse of the cumulative Euclidean distance to the equilibrium as an index of ecosystem resilience, further defined as $I_R$ (Fig. \@ref(fig:IrGraphFinal)):

@@

We tested the influence of three factors on the resilience index $I_R$: pre-disturbance species richness ($SR$), functional evenness ($FEve$) and functional diversity ($FDiv$) using linear regression ($I_R \sim SR + FEve + FDiv$). The analysis was conducted twice for each disturbance level (25%, 50% and 75% of basal area loss): one with forest structure properties and one with forest functioning properties. We tested variable correlation by Pearson's correlation coefficient for continuous variables and one-way ANOVA for species richness $SR$. In case of significant correlation, we selected the variable resulting in the model with the higher Aikaike information criterion in order to have independent variables. To meet the assumption of normality, the ecosystem resilience index $I_R$ was log-transformed.

For each ecosystem property $X$, we investigated the effect of biodiversity and its variation along post-disturbance trajectories. The net effect of biodiversity ($NE$) on ecosystem property $X$ is defined as the difference between the ecosystem property $X$ of the simulated mixed-species community and the one expected under the null hypothesis. The latter would mean that there is no effect of biodiversity, i.e. that ecosystem property $X$ for the mixed-species community equals the relative abundance-weighted sum of that in monoculture [@Loreau2001]. In order to do so, we simulated monoculture of each of the 163 species and for each level of disturbance leading to 652 simulations of monocultures.

@@

As species richness ($SR$) increased from 5 to 125 species, functional diversity varied from $0.76\pm0.18$ to $0.96\pm0.02$ and evenness ($FEve$) varied from $0.46\pm0.24$ to $0.64\pm0.02$. Species richness ($SR$) was significantly correlated to functional indices (one-way ANOVA $p-value < 0.001$ for both $FDiv$ and $FEve$) and model $AIC$ was higher without $SR$, consequently we kept only $FDiv$ and $FEve$ within linear regressions. Both functional diversity ($FDiv$) and functional evenness ($FEve$) were positively related with $I_R$, i.e. forest resilience increased with both functional diversity and functional evenness (Fig. \@ref(fig:gIr) and Table \@ref(tab:tgIr)). Similar results were obtained for all disturbance levels, with smaller p-values for higher disturbance intensity.

@@

The net effect of biodiversity on each ecosystem property was quantified and partitioned into selection and complementarity effects after normalizing, as explained in the Methods section. We found that the complementarity effect sharply increased immediately after disturbance, was the dominant effect of biodiversity during the first decades and then progressively vanished (Fig. \@ref(fig:gBE)). When increasing the disturbance intensity, the effect of complementarity increased further and dominated even longer. The selection effect, on the other hand, sharply decreased just after disturbance, and only gradually came to play a more important role, dominating the net biodiversity effect ca. a century after disturbance. Finally, at high disturbance intensity, the net effect of biodiversity on aboveground biomass was still lower than the control 600 years after disturbance (Fig. \@ref(fig:gBE)). We found similar results, but with an amplified signal, for basal area ($BA$) and stem abundance ($N$). Finally, forest growth primary productivity ($GPP$) recovered only within a few years with a dominant complementarity effect.

@@

Using an *in silico* experiment, we investigated the effect of biodiversity on tropical forest resilience to disturbance and the variation of biodiversity effect on ecosystem functioning along the recovery trajectory. We found that functional diversity improved tropical forest resilience from disturbance with underlying ecological mechanisms changing in dominance with time. Complementarity effect was dominant soon after disturbance and was progressively overtaken by a selection effect, as more competitive species dominated the forest community, a pattern that increased with disturbance intensity. Overall, our results suggest that pre-disturbance history is a key to our understanding of the ecological mechanisms that underlie the resilience of tropical forests. This can help reconciling previous contrasted results obtained with limited snapshots of ecosystem state.

@@

Our results provide strong evidence of a significant positive relationship between functional diversity and forest recovery rate. In particular, functional diversity, defined as the volume of the functional space occupied by the community [@villeger_new_2008], and its evenness component, both turned out to be major predictors of system resilience, as they helped avoid an overreliance on a few dominant species and the consequent vulnerability to their loss. This can be due to (i) temporal complementarity, with the diversity of species response to disturbances and response asynchronicity, resulting in ecological sorting [@Loreau2013; @Sakschewski2016; @Morin2014a], and (ii) functional dissimilarity reducing competition strength [@Loreau2013]. Our results thus confirm @Daz2001a, advocating against an under-evaluated importance of plant functional diversity in ecosystem processes, and @Zhang2012 who highlighted the role of evenness in productivity of forest ecosystems. In addition, increased species richness will *de facto* increase the chances for high functional diversity through the sampling effect [@Loreau1998; @Morin2014a].

@@

We found that complementarity among the species was the prevailing effect of biodiversity on forest functioning just after a disturbance. This is a direct consequence of the appearance of natural gaps and an increased heterogeneity in light availability, the main limiting resource in dense tropical forest. Species are known to vary widely in their demographic response to light availability depending on their functional strategy [@Herault2010; @Laurans2012; @Ruger2009]. Pioneer species with productive leaf tissue and light wood tend to occupy gaps fast after their opening, hence creating a shaded environment where late successional species with more conservative leaf tissue later establish. By inducing a higher diversity of shade tolerance and growth response to light in the species pool, species richness hence fosters forest response to perturbations [@Morin2011]. This complementarity effect decreased through time while the selection effect progressively gained in dominance. The diminished complementarity effect can be attributed to the increased abundance of the most competitive species in more stable and less heterogeneous environments.

@@

* METIST
* GFClim

>>>>

---
title: Biodiversity improves tropical forest resilience
author: Sylvain Schmitt$^{*,1,2}$, Isabelle Maréchaux$^3$, Jerome Chave $^4$, Fabian Jörg Fischer$^4$, Camille  Piponiot$^5$, Stéphane Traissac$^6$, Bruno Hérault$^{*,7,8}$
output:
  bookdown::word_document2: default
  bookdown::pdf_document2:
    includes:
      before_body: ./template/doc_prefix.tex
      in_header: ./template/preamble.tex
    number_sections: no
    toc: no
    keep_tex: yes
linestretch: 1.5
csl: /home/sylvain/Documents/Bibliography/csl/joe.csl
bibliography: /home/sylvain/Documents/Bibliography/library.bib
link-citations: yes
---

@@

$^3$ AMAP, INRA, Univ Montpellier, IRD, CIRAD, CNRS, F-34000 Montpellier, France

@@

$^5$ Université de Guyane, UMR EcoFoG (Agroparistech, CNRS, Inra, Université des
Antilles, Cirad), Campus Agronomique, 97310 Kourou, French Guiana

@@

$^7$ Cirad, Univ Montpellier, UR Forests & Societies, Montpellier, France

$^8$ INP-HB, Institut National Polytechnique Félix Houphouët-Boigny, BP 1093, Yamoussoukro, Ivory Coast

__$^*$ Corresponding authorw:__ sylvain.schmitt@agroparistech.fr ; bruno.herault@cirad.fr

@@

1. Climate change and direct human activities are modifying the disturbance regimes of tropical forests. Understanding the role of their high biological diversity in ecosystem recovery pathway and the underlying ecological mechanisms is of primary importance to predict the fate of tropical forests . Studies relying on permanent forest plots often lack repeatability, rarely include disturbed forests, and their temporal coverage remains too low to assess forest long-term dynamics. Forest simulators can thus prove useful to investigate the effect of biodiversity on forest functioning through virtual biodiversity experiments. However, the problem of biodiversity representation beyond a few plant functional types has long been an impediment in simulating species-rich tropical forests.
2. Using an individual-based and trait-based model of tropical forest, we assessed the effect of species and functional diversity on long-term forest recovery from disturbances of different levels by manipulating species number and functional assemblages across a large number of simulations. We further partitioned the net effect of biodiversity on ecosystem properties into complementarity and selection effects to investigate the ecological mechanisms that underlie the effect of biodiversity on forest functioning along recovery pathways.
3.  We found that functional diversity improved tropical forest resilience after disturbance. The ecological mechanisms underlying the effect of biodiversity on forest functioning changed in dominance along post-disturbance trajectories. The complementarity effect was dominant soon after disturbance and was progressively overtaken by a selection effect as more competitive species dominated the forest community, a pattern that increased with disturbance intensity.
4. *Synthesis*. We found that the mechanisms through which biodiversity influences forest functioning depend on the ecosystem state, shifting from a dominant complementarity effect in recently disturbed systems to a selection effect in anciently disturbed systems. Our results suggest that disturbance history is thus a key to our understanding of the biodiversity-ecosystem functioning relationships in tropical forests and can help reconcile contrasting results obtained previously with snapshots of ecosystem state in empirical studies.

@@

Resilience; Recovery; Functional diversity; Tropical forest; Complementarity effect; Selection effect; Biodiversity-ecosystem functioning.

@@

Climate change and direct human activities are currently deeply modifying forest disturbance regimes [@Davidson2012; @Lewis2015]. While deforestation has been the focus of worldwide attention, the consequences of human-induced disturbance on tropical forests have been less studied [@Putz2010; @Simula2009]. Degraded forests currently cover areas over ten times larger than deforested ones [@Herold2011], a majority of which is being unsustainably logged for timber production [@Blaser2011]. Forest degradation triggers severe biodiversity loss [@Burivalova2014] and is estimated to account for 2.1 billion of tons of carbon dioxide emissions per year representing 25% of greenhouse gas emissions related to deforestation and forest degradation globally [@Pearson2017].  In addition to direct human impact on forests, climate change modifies regional climates with, for instance, projected increases in the frequency and intensity of convective storms in the Amazon increasing wind-related tree mortality [@Negron-Juarez2018]. Understanding the drivers of tropical forest resilience to disturbances is thus critical. Here we explored the role of biodiversity in tropical forest recovery after disturbance through a simulation approach.

Ecological resilience refers to the ecosystem ability to resist and recover from disturbance [@Gunderson2010]. Post-disturbance recovery is determined by various factors, among which (i) the disturbance intensity which determines to a large extent the time to recover the ecosystem biomass, lost biodiversity, and tree species composition [@Burivalova2014; @DeAvila2015; @Rutishauser2015], (ii)  the pre-disturbance tree assemblage  as species respond differently to disturbances depending on their ecological strategies and functional properties [@herault_functional_2011] and (iii) tree community diversity which also fosters forest resilience to disturbances [@Cadotte2011; @Ives2007; @Loreau2013].

Two types of mechanisms can underlie the effect of biodiversity on ecosystem properties: complementarity and selection [@Loreau2001]. The complementarity effect results from resource partitioning and niche differentiation, and from facilitation, leading to a more efficient use of available resources at the community level. The selection effect describes how more diverse communities have a higher probability of including functionally efficient species. Moreover, the sampling effect implies redundancy and a lower risk to lose important functional traits, further improving resilience of species rich-forest [@Hooper2005a; @Tilman2001]. Previous theoretical work [@Loreau2010], experiments [@Cardinale2007; @Cardinale2009; @Niklaus2017a; @Huang2018], and direct observations along natural gradients [@Chisholm2013; @Liang2016; @Paquette2011; @Poorter2017] suggest that both complementarity and selection mechanisms may play a role in ecosystem biomass stock and dynamics [@VanderSande2017a], and their  relative importance can vary in time [@Fargione2007; @Huang2018].

@@

In the present study, we take advantage of recent developments made in the forest simulator TROLL [@Chave1999], an individual-based and spatially-explicit model of tropical forest that simulates high-diversity levels through a species-specific parameterization using plant functional traits [@Marechaux2017]. Consequently, TROLL offers the opportunity to study biodiversity-ecosystem functioning  relationships *in silico*. We used a simulation approach to assess the effect of species and functional diversity on long-term forest recovery from disturbances of different levels. We manipulated species number and functional identity across a large number of simulations, and measured the net effect of biodiversity on forest preperties along recovery pathways. To further investigate the underlying ecological mechanisms, we partitioned this net effect into complementarity and selection effects. Based on the general trend of a positive relationships between biodiversity and productivity [@Liang2016], we hypothesized that more diverse forests would recover quicker from a disturbance event due to an increased productivity.

@@

TROLL is an individual-based and spatially explicit forest dynamic model (details in @Marechaux2017). In TROLL, the life cycle (recruitment, growth, seed production, and death) is simulated for individual trees $>1\textrm{cm}$ diameter at breast height (dbh). Trees grow in a light environment computed within a $1\textrm{m}^3$ voxel grid. Seedlings below the $1\textrm{cm}$ size class are not modeled on an individual basis, but are part of a seed/seedlings pool. A maximum of one tree can establish per square meter. Once established, the tree is defined by age, dbh, height (h), crown radius (CR), crown depth (CD) and leaf area (LA). Tree geometry is calculated with allometric relationships. Additionally, each tree is flagged with a species label inherited from the parent tree through seedling recruitment. To each species label is associated a number of species-specific traits (Table \@ref(tab:traits)) from which ecophysiological processes are simulated using relationships drawn from the literature. The model thus represents tree biodiversity in a realistic way, where data requirement and model outputs parallel current trait collection efforts and field inventories.

Carbon assimilation is computed over half-hourly period of a representative day per month using the Farquhar, von Caemmerer, and Berry model of photosynthesis [@Farquhar1980].  Photosynthetic capacities, as well as leaf respiration rate, are computed from species-specific leaf concentration of nitrogen (N) and phosphorous (P), and leaf mass per area [LMA; @Atkin2015; @Domingues2010]. The net amount of assimilated carbon allocated to growth is then partitioned among the different plant organs using empirical factors. Yhe fraction allocated to woody growth is converted into increments of stem volume using species-specific wood density, while the fraction allocated to leaves is converted into leaf area increment using LMA. Leaf demography is simulated using species-specific leaf lifespan derived from LMA. Here we did not use the same relationship between LMA and leaf lifespan as in @Marechaux2017, as we found that it generated an underestimation of leaf lifespan for low LMA species. We thus designed a new leaf lifespan allometry, specific to tropical tree species, used throughout this study (see supplementary materials S1). Natural tree deaths are stochastic events and their rate is negatively correlated with wood density within TROLL, as observed pan-tropically [@kraft_functional_2010; @poorter_are_2008; @Wright2010] and locally [@Aubry-Kientz2013]. Additionally, trees can die due to carbon starvation or treefalls. Below-ground processes, herbaceous plants, epiphytes and lianas are not simulated. The source code is written in C++ and version 2.3.1 used in this study is available at https://github.com/TROLL-code/TROLL.

In order to have a general overview of disturbance effect, a generic disturbance module was developed for this study. At a given iteration $disturb_{iter}$, individual trees are randomly picked following a uniform law, targeting neither individuals nor species. This ensures that there are no confounding factors due to disturbance type specificity. Selected individuals are then removed without triggering a treefall to control disturbed biomass volume. The operation is repeated until the cumulative basal area of removed trees reaches the targeted threshold ($disturb_{intensity}$).

@@

In order to assess the role of biodiversity on ecosystem recovery from disturbance, we  simulated forest trajectories for different levels of disturbance intensity ($disturb_{intensity}$) and tree community compositions. For each simulation, we first conducted a spin-up run, where we simulated forest regeneration from bare soil for 600 years (pre-disturbance step). We then simulated the disturbance described as above ($disturb_{iter}$), and finally let the simulated forest community freely recover from the disturbance during 600 years (post-disturbance step). Disturbance intensity ($disturb_{intensity}$) was fixed at 0% (control), 25%, 50% or 75% of basal area loss. For each level of disturbance, we performed 20 simulations differing in the original number of species (species richness $SR\{15,25,125\}$) and in floristic composition. For each level of species richness, we first created 1000 virtual communities by randomly picking the desired number of species among the regional species pool. Then of those 1000 communities, we selected 20 communities with a large range of functional trait convex hull volume $CVH$ [@Cornwell2006a] but with a community mean ($CM$) close to the one of the regional species pool ($CM_{LMA} = 115 \pm 19 ~\textrm{g.m}^{-2}$, $CM_{wsg} = 0.73 \pm 0.07~\textrm{g.cm}^{-3}$, $CM_{dmax} = 0.37 \pm 0.1~\textrm{m}$, $CM_{hmax} = 40 \pm 11~\textrm{m}$) in order to control for average functional composition. This led to 240 simulations varying in community composition and disturbance intensity ($3~SR*20~CHV*4~ levels~of~disturbance$).

@@

where $X_D(t)$ and $X_C(t)$ are the values for the ecosystem property $X$ at time $t$ in the disturbed simulation ($disturb_{intensity} = 25\%, 50\%, 75\%$) and in the control simulation ($disturb_{intensity} = 0\%$), respectively. Note that variability among replicated simulations with the same set-up, due to demographic stochasticity in the model, is low enough to assume that a given simulation is representative of the community properties and dynamics given an initial species pool and disturbance intensity [@Marechaux2017].

When $R(t)$ approaches $R_{eq}=1$, the disturbed and control systems are considered equivalent, indicating a perfect recovery. We then calculated the Euclidean distance to this state of perfect recovery as $d_{eq}(t) = \sqrt{(R_{eq}-R(t))^2}$ (Fig. \@ref(fig:IrGraphFinal)). Because we set up the intensity of disturbance, our analysis did not include any ecosystem resistance, i.e. the ecosystem capacity to absorb a disturbance event. We used the inverse of the cumulative Euclidean distance to the equilibrium as an index of ecosystem resilience, further defined as $I_R$ (Fig. \@ref(fig:IrGraphFinal)):

@@

We tested the influence of three factors on the resilience index $I_R$: pre-disturbance species richness ($SR$), functional evenness ($FEve$) and functional diversity ($FDiv$):
$$log(I_R) \sim \mathcal{N} (\Theta _{SR} + \Theta _{FEve} \times Feve + \Theta _{FDiv} \times Fdiv; \sigma^2)$$
where $\Theta _{SR}$ is a vector of 3 parameters corresponding to  the 3 species richness levels, $\Theta _{FEve}$ and $\Theta _{FDiv}$ respectively model the effect of $FEve$ and $FDiv$ on $I_R$  under a normal law of variance $\sigma^2$. The analysis was conducted twice for each disturbance level (25%, 50% and 75% of basal area loss): one with forest structure properties and one with forest functioning properties. We used the AIC to select the final, most-parsimonious, models. The ecosystem resilience index IR was log-transformed to improve residuals’ normality.

For each ecosystem property $X$, we investigated the effect of biodiversity and its variation along post-disturbance trajectories. The net effect of biodiversity ($NE$) on ecosystem property $X$ is defined as the difference between the ecosystem property $X$ of the simulated mixed-species community and the one expected under the null hypothesis. The latter would mean that there is no effect of biodiversity, i.e. that ecosystem property $X$ for the mixed-species community equals the relative abundance-weighted sum of that in monoculture [@Loreau2001]. In order to do so, we simulated a monoculture of each of the 163 species and for each level of disturbance leading to 652 simulations of monocultures.

@@

As species richness ($SR$) increased from 5 to 125 species, functional diversity varied from $0.76\pm0.18$ to $0.96\pm0.02$ and evenness ($FEve$) varied from $0.46\pm0.24$ to $0.64\pm0.02$. Species richness ($SR$) was never retained by the AIC selection procedure, whatever the disturbance intensity. Both functional diversity ($FDiv$) and functional evenness ($FEve$) were positively related with $I_R$, i.e. forest resilience increased with both functional diversity and functional evenness (Fig. \@ref(fig:gIr) and Table \@ref(tab:tgIr)). Similar results were obtained for all disturbance levels, with higher significance for higher disturbance intensity. Model R2s were slightly lower for forest structure (from 0.18 to 0.27) than for forest functioning (from 0.27 to 0.51) variables. While FDiv was the most significant variable for forest structure models, Feve became prominent for forest functioning models (Table \@ref(tab:tgIr)).

@@

The net effect of biodiversity on each ecosystem property was quantified and partitioned into selection and complementarity effects (Table \@ref(tab:tNE)). The net effect was positive across most ecosystem variables, apart from 2, *i.e.* the number of stems above 10cm DBH and the net primary production for which the average net effects were negative, albeit not significantly different from 0. Regarding the aboveground biomass, we found that the complementarity effect sharply increased immediately after disturbance, was the dominant effect of biodiversity during the first decades and then progressively vanished (Fig. \@ref(fig:gBE)). When increasing the disturbance intensity, the effect of complementarity increased further and dominated even longer. The selection effect, on the other hand, sharply decreased just after disturbance, and only gradually came to play a more important role, dominating the net biodiversity effect ca. a century after disturbance. Finally, at high disturbance intensity, the net effect of biodiversity on aboveground biomass was still lower than the control 600 years after disturbance (Fig. \@ref(fig:gBE)). We found similar results, but with an amplified signal, for basal area ($BA$) and stem abundance ($N$). Finally, forest growth primary production ($GPP$) recovered only within a few years with a dominant complementarity effect.

@@

Using an *in silico* experiment, we investigated the effect of biodiversity on tropical forest resilience to disturbance and the variation of biodiversity effect on ecosystem functioning along the recovery trajectory. We found that functional diversity improved tropical forest recovery from disturbance with underlying ecological mechanisms changing in dominance with time. Complementarity effect was dominant soon after disturbance and was progressively overtaken by a selection effect, as more competitive species dominated the forest community, a pattern that increased with disturbance intensity. Overall, our results suggest that pre-disturbance history is key to our understanding of the ecological mechanisms that underlie the resilience of tropical forests. This can help reconciling previous contrasted results obtained with limited snapshots of ecosystem state.

@@

Our results provide strong evidence of a significant positive relationship between functional diversity and forest recovery rate. In particular, the volume of the functional space occupied by the community [@villeger_new_2008], and its evenness component, both turned out to be major predictors of system resilience, as they helped avoid an overreliance on a few dominant species and the consequent vulnerability to their loss. This can be due to (i) temporal complementarity, with the diversity of species response to disturbances and response asynchronicity, resulting in ecological sorting [@Loreau2013; @Sakschewski2016; @Morin2014a], and (ii) functional dissimilarity reducing competition strength [@Loreau2013]. Our results thus confirm @Daz2001a, advocating for an under-evaluated importance of plant functional diversity in ecosystem processes, and @Zhang2012 who highlighted the role of evenness in productivity of forest ecosystems. With our model simulations, we provide strong evidence that the high diversity of functional traits in tropical forests may represent an insurance against human-induced changes in disturbance regimes by maintaining a high ecosystem resilience.

@@

We found that complementarity among species was the prevailing effect of biodiversity on forest functioning just after a disturbance. This is a direct consequence of the appearance of natural gaps and an increased heterogeneity in light availability, the main limiting resource in dense tropical forest. Species are known to vary widely in their demographic response to light availability depending on their functional strategy [@Herault2010;@Ruger2009]. Pioneer species with productive leaf tissue and light wood tend to occupy gaps fast after their opening, hence creating a shaded environment where late successional species with more conservative leaf tissue later establish. By inducing a higher diversity of shade tolerance and growth response to light in the species pool, species richness hence fosters forest response to perturbations [@Morin2011]. This complementarity effect decreased through time while the selection effect progressively gained in dominance. The diminished complementarity effect can be attributed to the increased abundance of the most competitive species in more stable and less heterogeneous environments.

@@

This study was funded by the GFclim project (FEDER 20142020, Project GY0006894) and an Investissement d’Avenir grant of the ANR (CEBA: ANR-10-LABEX-0025).
